---
title: "Finding needle `conceptId`s"
author: "Jin Zhou"
date: "1 February 2022"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: 0
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(DT)
```

## Extracting needles from `covariate_ref` table

```{r initialize, include=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = keyring::key_get("legendT2dmDriverPath")) # NOTE to Jin: you need to set these key-chain variables for your computer

legendT2dmConnectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = paste(keyring::key_get("legendt2dmServer"),
                 keyring::key_get("legendt2dmDatabase"),
                 sep = "/"),
  user = keyring::key_get("legendt2dmUser"),
  password = keyring::key_get("legendt2dmPassword"))
```
  
```{r needles_covariate_ref}
con <- DatabaseConnector::connect(legendT2dmConnectionDetails)
covariates <- DatabaseConnector::querySql(con, sql = "SELECT DISTINCT covariate_id, covariate_name, concept_id FROM legendt2dm_class_diagnostics.covariate_ref;", snakeCaseToCamelCase = TRUE)
DatabaseConnector::disconnect(con)

covariates <- covariates %>% mutate(covariateName = toupper(covariateName))
matchNeedle <- grep(pattern = "NEEDLE", covariates$covariateName)
matchDevice <- grep(pattern = "DEVICE_EXPOSURE", covariates$covariateName)
matches <- intersect(matchNeedle, matchDevice)
needles <- covariates[matches,]
```

Table contains `r nrow(covariates)` distinct `covariateId`s.  

Of these, `r length(matches)` contain `DEVICE_EXPOSURE` and `NEEDLE`.

And, `r length(unique(needles$conceptId))` unique `concept_id` codes.

## Needle covariateIds from `covariate_ref` table`

```{r needles_covariate_ref_table}
datatable(needles, rownames = FALSE,
          options = list(order = list(2, 'desc')))
```
## Limiting to `PEN NEEDLE`

```{r needles_pen}
matchNeedle <- grep(pattern = "PEN NEEDLE", covariates$covariateName)
matchDevice <- grep(pattern = "DEVICE_EXPOSURE", covariates$covariateName)
matches <- intersect(matchNeedle, matchDevice)
penNeedles <- covariates[matches,]

datatable(penNeedles, rownames = FALSE,
          options = list(order = list(2, 'desc')))
```
## Limiting to `PEN NEEDLE` `conceptId`s

```{r pen_needle_concepts, echo=TRUE}
sort(unique(penNeedles$conceptId))
```

## Limiting to `NOVO NEEDLE`

```{r needles_novo}
matchNovo <- grep(pattern = "NOVO", covariates$covariateName)
matchNeedle <- grep(pattern = "NEEDLE", covariates$covariateName)
matchDevice <- grep(pattern = "DEVICE_EXPOSURE", covariates$covariateName)
matches <- intersect(matchNovo, intersect(matchNeedle, matchDevice))
novoNeedles <- covariates[matches,]

datatable(novoNeedles, rownames = FALSE,
          options = list(order = list(2, 'desc')))
```
## Limiting to `NOVO NEEDLE`

```{r novo_table, echo=TRUE}
sort(unique(novoNeedles$conceptId))
```

## Extracting needles from fitted PS models

```{r needles_ps_covariate_assessment}
con <- DatabaseConnector::connect(legendT2dmConnectionDetails)
psCovariates <- DatabaseConnector::querySql(con, sql = "
  SELECT DISTINCT ps_covariate_assessment.covariate_id, ps_covariate_assessment.covariate_name, concept_id 
  FROM legendt2dm_class_diagnostics.ps_covariate_assessment
  INNER JOIN legendt2dm_class_diagnostics.covariate_ref
  ON covariate_ref.covariate_id = ps_covariate_assessment.covariate_id;", snakeCaseToCamelCase = TRUE)
DatabaseConnector::disconnect(con)

psCovariates <- psCovariates %>% mutate(covariateName = toupper(covariateName))
psMatchNeedle <- grep(pattern = "NEEDLE", psCovariates$covariateName)
psNeedles <- psCovariates[psMatchNeedle,]
```

Fitted PS models contain `r nrow(psCovariates)` distinct `covariateId`s.  

Of these, `r length(psMatchNeedle)` contain `NEEDLE`.

And, `r length(unique(psNeedles$conceptId))` unique `concept_id` codes.

## Needle `covariateId` from fitted PS models

```{r needles_ps_covariate_assessment_table}
datatable(psNeedles, rownames = FALSE,
          options = list(order = list(2, 'desc')))
```

## Current fitted PS model `conceptId`s to exclude

```{r report, echo=TRUE}
sort(unique(psNeedles$conceptId))
```

## Current `conceptId`s to exclude

```{r report_full, echo=TRUE}
exclude <- sort(unique(c(penNeedles$conceptId,novoNeedles$conceptId)))
exclude
```

```{r, echo=TRUE}
paste(exclude, collapse = ";")
```

Limitations: would like to include VA results before modifying package
