---
title: "| RESEARCH PROTOCOL\n| \n| Large-scale evidence generation and evaluation
  across a network of databases for type 2 diabetes mellitus\n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    citation_package: natbib
    includes:
      before_body: title.tex
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: "style.css"
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '2020-09-24'
  version: 0.0.99
subtitle: 'Version: `r params$version`'
link-citations: true
csl: jamia.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
#knitr::knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage[table]{xcolor}', x, fixed = TRUE)})
library(dplyr)
```

# List of Abbreviations

```{r abbreviations, echo=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
  CDM; Common data model
  DPP4; Dipeptidyl peptidase-4 
  GLP1; Glucagon-like peptide-1
  IRB; Institutional review board
  LEGEND; Large-scale Evidence Generation and Evaluation across a Network of Databases
  MDRR; Minimum detectable risk ratio
  OMOP; Observational Medical Outcomes Partnership
  PS; Propensity score
  RCT; Randomized controlled trial
  SGLT2; Sodium-glucose co-transporter-2
  T2DM; Type 2 diabetes mellitus
")
kable(abbreviations, col.names = NULL, booktabs = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped")
```

# Responsible Parties

## Investigators

```{r parties, echo=FALSE}
parties <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Investigator; Institution/Affiliation
  George Hripcsak; Department of Biomedical Informatics, Columbia University, New York, NY, USA
  Harlan M. Krumholz; Department of Medicine, Yale University, New Haven, CT, USA
  Patrick B. Ryan; Observational Health Data Analytics, Janssen Research and Development, Titusville, NJ, USA
  Martijn J. Schuemie; Observational Health Data Analytics, Janssen Research and Development, Titusville, NJ, USA
  Marc A. Suchard *; Department of Biostatistics, University of California, Los Angeles, Los Angeles, CA, USA
")
kable(parties, booktabs = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped") %>%
  column_spec(1, width = "10em") %>%
  column_spec(2, width = "30em") %>%
  footnote(general = "* Principal Investigator", general_title = "")
```

## Disclosures

This study is undertaken within Observational Health Data Sciences and Informatics (OHDSI), an open collaboration.
**GH** receives grant funding from the US National Institutes of Health and the US Food & Drug Administration and contracts from Janssen Research and Development.
**PBR** and **MJS** are employees of Janssen Research and Development and shareholders in John & Johnson.
**MAS** receives grant funding from the US National Institutes of Health and the US Food & Drug Administration and contracts from Janssen Research and Development and IQVIA.

# Abstract

When in the Course of human events, it becomes necessary ...

# Amendments and Updates

```{r amendments, echo=FALSE}
amendments <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Number; Date; Section of study protocol; Amendment or update; Reason
  None;;;; 
")
options(knitr.kable.NA = "")
kable(amendments, booktabs = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped")
```

```{r dates, echo=FALSE}
dates <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Milestone; Planned / actual date
  EU PAS Registration; 
  Start of analysis;
  End of analysis;
  Results presentation;
")
options(knitr.kable.NA = "")
kable(dates, booktabs = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped")
```


# Rationale and Background

The landscape of therapeutic options for type 2 diabetes mellitus (T2DM) has been dramatically transformed over the last decade [@Lo2018-ca].
The emergence of drugs targeting the sodium-glucose co-transporter-2 (SGLT2) and the glucagon-like peptide-1 (GLP1) receptor has expanded the role of T2DM agents from lowering blood glucose to directly reducing cardiovascular risk [@North2019-xn].
A series of large randomized clinical trials designed to evaluate the cardiovascular safety of SGLT2 inhibitors and GLP1 receptor agonists found that use of many of these agents led to a reduction in major adverse cardiovascular events, including myocardial infarction, hospitalization for heart failure, and cardiovascular mortality [@Zinman2015-ps; @Neal2017-gs; @Marso2016-zg; @Marso2016-yk].
However, other T2DM drugs widely used before the introduction of these novel agents, such as sulfonylureas, did not undergo similarly comprehensive trials to evaluate their cardiovascular efficacy or safety.
Moreover, direct comparisons of newer agents with dipeptidyl peptidase-4 (DPP4) inhibitors, with neutral effects on major cardiovascular outcomes [@Scirica2013-gl; @White2013-zl; @Green2015-pv; @Rosenstock2019-cp], have not been conducted. 
Nevertheless, DPP4 inhibitors and sulfonylureas continue to be used in clinical practice and are recommended as second line T2DM agents in national clinical practice guidelines.

Several challenges remain in formulating T2DM treatment recommendations based on existing evidence [@Cefalu2018-mu].
First, trials of novel agents did not pursue head-to-head comparisons to older agents and were instead designed as additive treatments on the background of commonly used T2DM agents.
Therefore, the relative cardiovascular efficacy and safety of novel compared with older agents is not known.
Second, trials of novel agents have tested individual drugs against placebo, but have not directly compared SGLT2 inhibitors with GLP1 receptor agonists in reducing adverse cardiovascular event risk.
Moreover, there is no evidence to guide the use of individual drugs within each class and across different drug classes, particularly among patients at lower cardiovascular risk than recruited in clinical trials.
Third, randomized trials focused on cardiovascular efficacy and safety, but were not powered to adequately assess the safety of these agents across a spectrum of non-cardiovascular outcomes.
Finally, restricted enrollment across regions, and subgroups of age, sex, and race further limits the efficacy and safety assessment that may guide individual patients’ treatment.

Evidence gaps from these trials also pose a challenge in designing treatment algorithms, which rely on comparative effectiveness and safety of drugs.
Perhaps, as a result, there is large variation in clinical practice guidelines and in clinical practice with regard to these medications, with many patients initiated on the newer therapies and many others treated with older regimens [@McCoy2019-hu; @Curtis2018-zg; @Arnold2017-fo; @Dave2020-ql; @Le2020-lh].
Among the second-line options, there is much variation with respect to the order of drugs used.
This lack of consensus about the best approach provides an opportunity for systematic, large-scale observational studies. 

# Study Objectives

To inform critical decisions facing patients with diabetes, their caregivers, clinicians, policymakers and healthcare system leaders, we have launched the Large-Scale Evidence Generation and Evaluation across a Network of Databases for Diabetes (LEGEND-T2DM) initiative to execute a series of comprehensive observational studies to compare cardiovascular outcome rates and safety of second-line T2DM glucose-lowering agents. Specifically, these studies aim

1. To determine, through systematic evaluation, the comparative effectiveness of second-line T2DM agents, SGLT2 inhibitors and GLP1 receptor agonists, with each other and with DPP4 inhibitors and sulfonylureas, for cardiovascular outcomes.
2. To determine, through systematic evaluation, the comparative safety of second-line T2DM agents among patients with T2DM.
3. To assess heterogeneity in effectiveness and safety of second-line T2DM agents among key patient subgroups: Using stratified patient cohorts, we will quantify differential effectiveness and safety across subgroups of patients based on age, sex, race, renal impairment, and baseline cardiovascular risk.  

# Research Methods

LEGEND-T2DM will execute three systematic, large-scale observational studies of second-line T2DM agents to estimate the relative risks of cardiovascular effectiveness and safety outcomes.

1. The **Class-vs-Class Study** will provide all pairwise comparisons between the four major T2DM agent classes to evaluate their comparative effects on cardiovascular risk  (Aim 1) and patient-centered safety outcomes (Aim 2);  
2. The **Drug-vs-Drug Study** will furnish head-to-head pairwise comparisons between individual agents within and across classes (both Aims 1 and 2); and 
3. The **Heterogeneity Study** will refine these comparisons for T2DM patients for important subgroups (Aim 3).   
In contrast to a single comparison approach, LEGEND-T2DM will provide a comprehensive view of the findings and their consistency across populations, drugs, and outcomes.
We will model each study on our successful collaborative research evaluating the comparative effectiveness of antihypertensives recently published in _The Lancet_ [@Suchard2019-gq].

For each study, we will employ an active comparator, new-user cohort design [@Yoshida2015-ln; @Ryan2013-wq; @Schuemie2020-wx].
New-user cohort design is advocated as the primary design to be considered for comparative effectiveness and drug safety [@Schneeweiss2010-my; @Gagne2012-at; @Johnson2013-db].
By identifying patients who start a new treatment course and using therapy initiation as the start of follow-up, the new-user design models an randomized controlled trial (RCT) where treatment commences at the index study visit.
Exploiting such an index date allows a clear separation of baseline patient characteristics that occur prior to index date and are usable as covariates in the analysis without concern of inadvertently introducing mediator variables that arise between exposure and outcome [@Schneeweiss2007-am]. 
Excluding prevalent users as those without a sufficient washout period prior to first exposure occurrence further reduces bias due to balancing mediators on the causal pathway, time-varying hazards, and depletion of susceptibles.
[@Johnson2013-db; @Suissa2017-lv]. 
Our systematic framework across studies in this proposal further will address residual confounding, publication bias, and p-hacking using data-driven, large-scale propensity adjustment for measured confounding [@Tian2018-xy], a large set of negative control outcome experiments to address unmeasured and systematic bias
[@Schuemie2014-bv; @Schuemie2016-km; @Schuemie2018-hq], and full disclosure of hypotheses tested [@Schuemie2018-zi].

## Data Sources

We will execute LEGEND-T2DM as a series of OHDSI network studies.
All data partners within OHDSI are encouraged to participate voluntarily and can do so conveniently, because of the community's shared Observational Medical Outcomes Partnership (OMOP) common data model (CDM) and OHDSI tool-stack.
Many OHDSI community data partners have already committed to participate and we will recruit further data partners through OHDSI’s standard recruitment process, which includes protocol publication on OHDSI’s GitHub, an announcement in OHDSI’s research forum, presentation at the weekly OHDSI all-hands-on meeting and direct requests to data holders.

Table \@ref(tab:data-sources) lists the 13 already committed data sources for LEGEND-T2DM; these sources encompass a large variety of practice types and populations. 
For each data source, we report a brief description and size of the population it represents and its patient capture process and start date.
While the earliest patient capture begins in 1989 (CUIMC), the vast majority come from the mid-2000s to today, providing almost two decades of T2DM treatment coverage.
US populations include those commercially and publicly insured, enriched for older individuals (MDCR, VA), lower socioeconomic status (MDCD), and racially diverse (VA >20% Black or African American, CUIMC 8%).
The US data sources may capture the same patients across multiple sources.
Different views of the same patients are an advantage in capturing the diversity of real-world health events that patients experience.
Across CCAE (commercially insured), MCDR (Medicare)  and MCDC (Medicaid), we expect little overlap in terms of the same observations recorded at the same time for a patient; patients can flow between sources (e.g., a CCAE patient who retires can opt-in to become an MDCR patient), but the enrollment time periods stand distinct.
On the other hand, Optum, PanTher, OpenClaims, CUIMC and YNHHS may overlap in time with the other US data sources.
While it remains against licensing agreements to attempt to link patients between most data sources, Optum reports <20% overlap between their claims and EHR data sources that is reassuringly small.
All data sources will receive institutional review board approval or exemption for their participation before executing LEGEND-T2DM.

```{r data-sources, echo=FALSE, warning=FALSE}
data_sources <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Data source ; Population ; Patients ; History ; Data capture process and short description
  IBM MarketScan Commercial Claims and Encounters (CCAE) ; Commercially insured, < 65 years ; 142M ; 2000 -- ; Adjudicated health insurance claims (e.g. inpatient, outpatient, and outpatient pharmacy)  from large employers and health plans who provide private healthcare coverage to employees, their spouses and dependents.
  IBM MarketScan Medicare Supplemental Database (MDCR)  ; Commercially insured, 65$+$ years ; 10M ; 2000 -- ; Adjudicated health insurance claims of retirees with primary or Medicare supplemental coverage through privately insured fee-for-service, point-of-service or capitated health plans.
  IBM MarketScan Multi-State Medicaid Database (MDCD) ; Medicaid enrollees, racially diverse ; 26M ; 2006 -- ; Adjudicated health insurance claims for Medicaid enrollees from multiple states and includes hospital discharge diagnoses, outpatient diagnoses and procedures, and outpatient pharmacy claims.
  IQVIA Open Claims (IOC) ; General ; 160M ; 2010 -- ; Pre-adjudicated claims at the anonymized patient-level collected from office-based physicians and specialists via office management software and clearinghouse switch sources for the purpose of reimbursement.
  Japan Medical Data Center (JMDC) ; Japan, general ; 5.5M ; 2005 -- ; Data from 60 society-managed health insurance plans covering workers aged 18 to 65 and their dependents.
  Korea National Health Insurance Service (NHIS) ; 2% random sample of South Korea ; 1M ; 2002 -- ; National administrative claims database covering the South Korean population.
  Optum Clinformatics Data Mart (Optum) ; Commercially or Medicare insured ; 85M ; 2000 -- ; Inpatient and outpatient healthcare insurance claims.
  Columbia University Irving Medical Center (CIUMC) ; Academic medical center patients, racially diverse ; 6M ; 1989 -- ; General practice, specialists and inpatient hospital services from the New York-Presbyterian hospital and affiliated academic physician practices in New York.
  Department of Veterans Affairs (VA) ; Veterans, older, racially diverse ; 12M ; 2000 -- ; National VA health care system, the largest integrated provider of medical services in the US, provided at 170 VA medical centers and 1,063 outpatient sites.
  Information System for Research in Primary Care (SIDIAP) ; 80% of all Catalonia (Spain) ; 7.7M ; 2006 -- ; Primary care partially linked to inpatient data with pharmacy dispensations and primary care laboratories.  Healthcare is universal and taxpayer funded in the region, and PCPs are gatekeeps for all care and responsible for repeat prescriptions. 
  IQVIA Disease Analyzer Germany (DAG) ; Germany, general ; 37M ; 1992 -- ; Collection from patient management software used by general practitioners and selected specialists to document patients’ medical records within their office-based practice during a visit.
  Optum Electronic Health Records (OptumEHR) ; US, general ; 93M ; 2006 -- ; Clinical information, prescriptions, lab results, vital signs, body measurements, diagnoses and procedures derived from clinical notes using natural language processing. 
  Yale New Haven Health System (YNHHS) ; Academic medical center patients ; 2M ; 2013 -- ; General practice, specialists and inpatient hospital services from the YNHHS in Connecticut.
")
tab <- kable(data_sources, booktabs = TRUE,
      caption = "Committed LEGEND-T2DM data sources and the populations they cover.") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped") %>%
  pack_rows("Administrative claims", 1, 7, latex_align = "c", indent = FALSE) %>%
  pack_rows("Electronic health records (EHRs)", 8, 13, latex_align = "c", indent = FALSE)

if (knitr::is_latex_output()) {
  tab %>%
    column_spec(1, width = "10em") %>%
    column_spec(2, width = "10em") %>%
    column_spec(5, width = "25em") %>%
    kable_styling(font_size = 8)
} else {
  tab
}
```










## Patient Cohorts

## Exposures

## Controls or Comparators

## Outcomes

```{r outcomes, echo=FALSE, warning=FALSE, results='asis'}
outcomes <- read.csv("OutcomeCohortTable.csv")

if (knitr::is_latex_output()) {
  convertCitationToLatex <- function(cite) {
    if (cite != "") {
      tmp <- substring(cite, 2, nchar(cite) - 1)
      citeList <- strsplit(tmp, ";")
      tmpList <- lapply(citeList, function(x) {
        substring(x, 2, nchar(x))
      })
      tags <- paste(unlist(tmpList), collapse = ",")
      cite <- paste("\\cite{", tags, "}", sep = "")
    }
    cite
  }
  outcomes$cite <- sapply(outcomes$cite, convertCitationToLatex)
}

tab <- kable(outcomes, booktabs = TRUE, escape = FALSE,
      caption = "Committed LEGEND-T2DM data sources and the populations they cover.") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped") 

if (knitr::is_latex_output()) {
  tab %>%
    column_spec(3, width = "25em") %>%
    column_spec(4, width = "10em") %>%
    kable_styling(font_size = 8)
} else {
  tab
}
```

### TODO: Additional outcomes

* Which additional outcomes from the table below from LEGEND-HTN are also relevant for T2DM?

```{r additional-oucomes, echo=FALSE, warning=FALSE}
htn_outcomes <- read.csv("LegendHtnTable.csv")
tab <- kable(htn_outcomes %>% select(name), booktabs = TRUE,
      caption = "All LEGEND-HTN outcomes") %>%
  kable_styling(bootstrap_options = "striped", latex_options = "striped")
if (knitr::is_latex_output()) {
  tab %>%
    kable_styling(font_size = 8)
} else {
  tab
}
```


## Covariates

## Analysis

# Sample Size and Study Power

Within each data source, we will execute all comparisons with $\ge$ 1,000 eligible patients per arm.
Blinded to effect estimates, investigators and stakeholders will evaluate extensive study diagnostics for each comparison to assess reliability and generalizability, and only report risk estimates that pass [@Schuemie2018-zi; @Schuemie2020-he].
These diagnostics will include 

1. Minimum detectable risk ratio (MDRR) as a typical proxy for power, 
2. Preference score distributions to evaluate empirical equipoise10 and population generalizability, 
3. Extensive patient characteristics to evaluate cohort balance before and after PS-adjustment, 
4. Negative control calibration plots to assess residual bias, and 
5. Kaplan-Meier plots to examine hazard ratio proportionality assumptions.

We will define cohorts to stand in empirical equipoise if the majority of patients carry preference scores between 0.3 and 0.7 and to achieve balance if all after-adjustment characteristics return absolute standardized mean differences $<$ 0.1 [@Austin2009-af].

### TODO: Meta-analysis heterogeneity measures $i^2$. How does this apply to our total evidence approach?

# Strengths and Limitations

## Strengths
- Cohort studies allow direct estimation of incidence rates following exposure of interest, and the new-user design can capture early events following treatment exposures while avoiding confounding from previous treatment effects; new use allows for a clear exposure index date.
- Large-scale propensity score matching and stratification create balance on a large number of baseline potential confounders and have been found in the past to balance unmeasured confounders.
- Systematic processes including a pre-specified selection of covariates avoids investigator-specific biases in variable selection.
- Use of real negative and synthetic positive control outcomes provides an independent estimate of residual bias in the experiment.
- The fully specified study protocol is being published before analysis begins.
- Dissemination of the results will not depend on estimated effects, avoiding publication bias.
- All analytic methods have previously been verified on real data.
- All software is freely available as open source.
- Use of a common data model allows extension of the experiment to future databases and allows replication of these results on licensable databases that were used in this experiment, while still maintaining patient privacy on patient-level data.
- Use of multiple databases allows estimating consistency to add credibility and supports generalizability.
 
## Limitations
 
- Even though many potential confounders will be included in this study, there may be residual bias due to unmeasured or misspecified confounders, such as confounding by indication, differences in physician characteristics that may be associated with drug choice, concomitant use of other drugs started after the index date, and informative censoring at the end of the on-treatment periods. To minimize this risk, we used methods to detect residual bias through our negative and positive controls.
- Our follow-up times are limited and variable, potentially reducing power to detect differences in effectiveness and safety.
- We assume hazards are not time varying.
- Misclassification of study variables is unavoidable in secondary use of health data, so it is possible to misclassify treatments, covariates, and outcomes; we do not expect differential misclassification, so bias will most likely be towards the null.
- The electronic health record databases may be missing care episodes for patients due to care outside the respective health systems; bias will most likely be towards the null.

# Protection of Human Subjects

LEGEND-T2DM does not involve human subjects research.
The project does, however, use de-identified human data collected during routine healthcare provision.
All data partners executing the LEGEND-T2DM studies within their data sources will have received institutional review board (IRB) approval or waiver for participation in accordance to their institutional governance prior to execution (see Table ADD-REF).
LEGEND-T2DM executes across a federated and distributed data network, where analysis code is sent to participating data partners and only aggregate summary statistics are returned, with no sharing of patient-level data between organizations.

```{r irb, echo=FALSE}
data_sources <- readr::read_delim(col_names = TRUE, delim = "&", trim_ws = TRUE, file = "
Data source & Statement
IBM MarketScan Commercial Claims and Encounters (CCAE) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
IBM MarketScan Medicare Supplemental Database (MDCR)  & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
IBM MarketScan Multi-State Medicaid Database (MDCD) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
IQVIA Open Claims (IOC) & This is a retrospective database study on de-identified data and is deemed not human subject research. Approval is provided for OHDSI network studies.
Japan Medical Data Center (JMDC) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
Korea National Health Insurance Service (NHIS) & Ajou University Institutional Review Board (AJIRB-MED-EXP-17-054 for LEGEND-HTN) and approval expected shortly for LEGEND-T2DM.
Optum Clinformatics Data Mart (Optum) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
Columbia University Irving Medical Center (CIUMC) & Use of the CUIMC data source was approved by the Columbia University Institutional Review Board as an OHDSI network study (IRB\\# AAAO7805).
Department of Veterans Affairs (VA) & Use of the VA-OMOP data source was reviewed by the Department of Veterans Affairs Central Institutional Review Board (IRB) and was determined to meet the criteria for exemption under Exemption Category 4(3) and approved the request for Waiver of HIPAA Authorization.
Information System for Research in Primary Care (SIDIAP) & Use of the SIDIAP data source was approved by the Clinical Research Ethics Committee of IDIAPJGol (project code: 20/070-PCV)
IQVIA Disease Analyzer Germany (DAG) & This is a retrospective database study on de-identified data and is deemed not human subject research. Approval is provided for OHDSI network studies.
Optum Electronic Health Records (OptumEHR) & New England Institutional Review Board and was determined to be exempt from broad IRB approval, as this research project did not involve human subject research.
Yale New Haven Health System (YNHHS) & Use of the YNHHS EHR data source was approved by the Yale University Institutional Review Board as an OHDSI network study (IRB\\# pending).
")

tab <- kable(data_sources, booktabs = TRUE,
      caption = "IRB approval or waiver statement from partners.") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "striped")

if (knitr::is_latex_output()) {
  tab %>%
    column_spec(1, width = "15em") %>%
    column_spec(2, width = "40em") %>%
    kable_styling(font_size = 8)
} else {
  tab
}
```

# Management and Reporting of Adverse Events and Adverse Reactions

LEGEND-T2DM uses coded data that already exist in electronic databases.
In these types of databases, it is not possible to link (i.e., identify a potential causal association between) a particular product and medical event for any specific individual.
Thus, the minimum criteria for reporting an adverse event (i.e., identifiable patient, identifiable reporter, a suspect product and event) are not available and adverse events are not reportable as individual adverse event reports.
The study results will be assessed for medically important findings.

# Plans for Disseminating and Communicating Study Results

Open science aims to make scientific research, including its data process and software, and its dissemination, through publication and presentation, accessible to all levels of an inquiring society, amateur or professional [@Woelfle2011-ss] and is a governing principle of LEGEND-T2DM.
Open science delivers reproducible, transparent and reliable evidence.
All aspects of LEGEND-T2DM (except private patient data) will be open and we will actively encourage other interested researchers, clinicians and patients to participate.
This differs fundamentally from traditional studies that rarely open their analytic tools or share all result artifacts, and inform the community about hard-to-verify conclusions at completion.

## Transparent and re-usable research tools

We will publicly register this protocol and announce its availability for feedback from stakeholders, the OHDSI community and within clinical professional societies.
This protocol will link to open source code for all steps to generating diagnostics, effect estimates, figures and tables.
Such transparency is possible because we will construct our studies on top of the OHDSI toolstack of open source software tools that are community developed and rigorously tested [@Schuemie2020-wx].
We will publicly host LEGEND-T2DM source code at (https://github.com/ohdsi-studies/LegendT2dm), allowing public contribution and review, and free re-use for anyone’s future research.

## Continous sharing of results

LEGEND-T2DM embodies a new approach to generating evidence from healthcare data that overcome weaknesses in the current process of answering and publishing (or not) one question at a time.
Generating evidence for thousands of research and control questions using a systematic process enables us to not only evaluate that process and the coherence and consistency of the evidence, but also to avoid $p$-hacking and publication bias [@Schuemie2018-zi].
We will store and openly communicate all of these results as they become available using a user-friendly web-based app that serves up all descriptive statistics, study diagnostics and effect estimates for each cohort comparison and outcome.
Open access to this app will be through a general public facing LEGEND-T2DM web-page.

## Scientific meetings and publications

We will deliver multiple presentations annually in scientific venues including the annual meetings of the American Diabetes Association, American College of Cardiology, American Heart Association and American Medical Informatics Association.
We will also prepare multiple scientific publications for clinical, informatics and statistical journals.

## General public

We believe in sharing our findings that will guide clinical care with the general public.
LEGEND-T2DM will use social-media (Twitter) to facilitate this.
With dedicated support from the OHDSI communications specialist, we will deliver regular press releases at key project stages, distributed via the extensive media networks of UCLA, Columbia and Yale.

# Lists of Tables and Figures

# References

<div id="refs"></div>

# Appendix 1: Literature Review

# Appendix 2: Target, Comparator and Outcome Cohort Definitions

## Acute myocardial infarction

```{r test, echo=FALSE, results="asis", cache=TRUE}
cohort <- ROhdsiWebApi::getCohortDefinition(cohortId = 1247, baseUrl = "http://atlas-covid19.ohdsi.org/WebAPI/")
expression <- RJSONIO::toJSON(cohort$expression)
markdown <- CirceR::cohortPrintFriendly(expression)
writeLines(markdown)
```
