---
title: "Propensity Score Model Assessment"
author: "Marc A. Suchard"
date: "23 June 2021"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
    transition: 0
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(DT)
```

## Class-vs-class data experiment

Available data sources:

- CCAE
- MDCR 
- OptumEHR (_only_ main and female-stratum)

Results are *now* available on `legend_t2dm_class_diagnostics`

RMarkdown file `PsModelAssessment.Rmd` (in `LegendT2dm` repo) generates this document using the OHDSI public DB server

```{r load_from_database, include=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = keyring::key_get("legendT2dmDriverPath"))

legendT2dmConnectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = paste(keyring::key_get("legendt2dmServer"),
                 keyring::key_get("legendt2dmDatabase"),
                 sep = "/"),
  user = keyring::key_get("legendt2dmUser"),
  password = keyring::key_get("legendt2dmPassword"))

con <- DatabaseConnector::connect(legendT2dmConnectionDetails)
auc <- DatabaseConnector::querySql(con, sql = "SELECT * FROM legendt2dm_class_diagnostics.ps_auc_assessment;", snakeCaseToCamelCase = TRUE)
ps <- DatabaseConnector::querySql(con, sql = "SELECT * FROM legendt2dm_class_diagnostics.ps_covariate_assessment;", snakeCaseToCamelCase = TRUE)
DatabaseConnector::disconnect(con)

auc <- auc %>% mutate(auc = sprintf("%1.4f", auc),
                      equipoise = sprintf("%1.2f", equipoise)) %>%
  select(auc, equipoise, comparison, databaseId)

ps <- ps %>% mutate(sign = ifelse(coefficient >= 0, "+", "-"),
                    absCoef = sprintf("%1.2f", abs(coefficient))) %>%
  select(absCoef, covariateName, comparison, databaseId) 

```
  
## Propensity score AUC and equipoise

AUC close to (but not less than) 0.5 is best

Equipoise computed for [0.3 - 0.7] interval
  
```{r auc}
datatable(auc, rownames = FALSE,
          options = list(order = list(2, 'desc')))
```

## Most separable covariates

Note: `index year: 2013`, `BD PEN NEEDLE` and `INSULIN AND ANALOGUES`

```{r covariates}
datatable(ps, rownames = FALSE,
          options = list(order = list(2, 'desc'),
                         pageLength = 5))
```
