---
title: "| OPERATIONS MANUAL\n| \n| Large-scale evidence generation and evaluation
  across a network of databases for type 2 diabetes mellitus\n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    css: "../Protocol/style.css"
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    # citation_package: natbib
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{../Protocol/header.tex}
longtable: yes
mainfont: Arial
bibliography: ../Protocol/Protocol.bib
params:
  date: '2020-09-24'
  version: 0.0.99
subtitle: 'Version: `r params$version`'
link-citations: true
csl: ../Protocol/jamia.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
#knitr::knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage[table]{xcolor}', x, fixed = TRUE)})
library(dplyr)
options(knitr.kable.NA = "")

# pdf2png <- function(path) {
#   # only do the conversion for non-LaTeX output
#   if (knitr::is_latex_output()) {
#     return(path)
#   }
#   path2 <- xfun::with_ext(path, "png")
#   img <- magick::image_read_pdf(path)
#   magick::image_write(img, path2, format = "png")
#   path2
# }

latex_table_font_size <- 8
```

# TODO List {-}

  * Fix export of ps_assessment to remove NULL coefficients
  * Make corresponding SQL column NOT NULL again
  * IIT end of obs period for VA
  * HbA1c control (remove earliest event and remove subject prior to outcome)

<!-- * item -->

# List of Abbreviations

```{r abbreviations, echo=FALSE}
abbreviations <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
  CDM; Common data model
  DPP4; Dipeptidyl peptidase-4
  GLP1; Glucagon-like peptide-1
  IRB; Institutional review board
  LEGEND; Large-scale Evidence Generation and Evaluation across a Network of Databases
  MACE; Major adverse cardiovascular event
  MDRR; Minimum detectable risk ratio
  OHDSI; Observational Health Data Science and Informatics
  OMOP; Observational Medical Outcomes Partnership
  PS; Propensity score
  RCT; Randomized controlled trial
  SGLT2; Sodium-glucose co-transporter-2
  T2DM; Type 2 diabetes mellitus
")

tab <- kable(abbreviations, col.names = NULL, linesep = "", booktabs = TRUE)

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Getting Started

LEGEND-T2DM consists of several `R` packages available on https://github.com/ohdsi-studies:

- `LegendT2dm`
- `LegendT2dmCohortExplorer` and
- `LegendT2dmEvidenceExplorer`

that depend themselves on a large number of packages from Hades (add link).
To gracefully handle these dependencies, LEGEND-T2DM uses `renv` through a version-controlled `renv.lock` file to specify each package.
To install `LegendT2dm` and all of its dependencies, we first "pull" the lastest source from `github`:

```
git clone https://github.com/ohdsi-studies/LegendT2dm
```

then from within an `R` or `Rstudio` terms, we active `renv`:

```{r, eval=FALSE}
install.packages("renv") # Only necessary once
renv::activate()
renv::restore()
```

after which one can now build and install the final `LegendT2dm` package.

# Exposure Cohort Counts

For the **Class-vs-Class Study**, we will construct
$$
\begin{aligned}
& (4 \text{ drug-class targets})\ \times \\
& (2 \text{ On-treatment censoring types})\ \times \\
& (2 \text{ Prior metformin-use choices}) \\
& = 16
\end{aligned}
$$
exposure cohort definitions.
Likewise, for the **Drug-vs-Drug Study**, we will construct
$$
\begin{aligned}
& (22 \text{ drug-ingredient targets})\ \times \\
& (2 \text{ On-treatment censoring types})\ \times \\
& (2 \text{ Prior metformin-use choices}) \\
& = 88
\end{aligned}
$$
exposure cohort definitions.
For the **Heterogeneity Study**, we will stratify each of these cohort definitions into
$$
\begin{aligned}
& 3 \text{ (age: 18-44 / 45-64 /}\ge\text{65) } + \\
& 2 \text{ (sex: female / male) } + \\
& 1 \text{ (race: Black) } + \\
& 2 \text{ (cardiovascular risk: lower / higher) } + \\
& 2 \text{ (renal impairment: with/without) } \\
& = 10
\end{aligned}
$$
additional definitions, results in $(16 + 88) \times 10 = 1040$ further cohort definitions.

# Exposure Cohort ID Key

This section documents internal coding used across the three studies to label cohort definitions and their SQL representation.  Each cohort carries 9-digit identifier
$$ \text{ID}\ [DD][T][M][A][S][R][C][K] $$
where Table \@ref(tab:drug-codes) reports the target 2-digit drug codes $DD$ and Table \@ref(tab:category-codes) lists the codes for time-at-risk $T$, metformin $M$, age $A$, sex $S$, race $R$, cardiovascular $C$ and renal (kidney) $K$ categories.

```{r drug-codes, echo=FALSE, message=FALSE, warning=FALSE}
addTab <- function(name, cohortId) {
  if (cohortId %% 10 != 0) {
    paste0("\\ \\ \\ \\ ", name)
  } else {
    name
  }
}

exposures <- read.csv(system.file("settings/ExposuresOfInterest.csv", package = "LegendT2dm"))

drugCodes <- exposures  %>% arrange(cohortId) %>%
  rowwise() %>%
  mutate(name = addTab(name, cohortId)) %>%
  select(name, cohortId) %>%
  rename(Drug = name, "2-digit Code" = cohortId)

tab <- kable(drugCodes, booktabs = TRUE, linesep = "",
             caption = "2-Digit Drug Codes") %>%
  column_spec(1, width = "10em") %>%
  column_spec(2, width = "30em")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

```{r category-codes, echo=FALSE, warning=FALSE, message=FALSE}
categoryCodes <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  Category; Value; 1-digit Code
  Time-at-risk;
  ; ITT/OT1; 1
  ; OT2; 2
  Metformin;
  ; prior; 1
  ; none; 2
  Age ;
  ; any; 0
  ; younger; 1
  ; middle; 2
  ; older; 3
  Sex ;
  ; any; 0
  ; female; 1
  ; male; 2
  Race ;
  ; any; 0
  ; Black; 1
  Cardiovascular risk ;
  ; any; 0
  ; low; 1
  ; higher; 2
  Renal disease ;
  ; any; 0
  ; without; 1
  ; with; 2
")

footnote <- c("ITT: intent-to-treat; OT1: on-treatment-1; OT2: on-treatment-2")

tab <- kable(categoryCodes, booktabs = TRUE, linesep = "",
             caption = "Other Cohort Category Codes") %>%
  footnote(general = footnote, general_title = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

Here are useful `regex` masks for the different strata

```{r, echo=FALSE, message=FALSE}
mask <- readr::read_csv(system.file("settings", "masks.csv", package = "LegendT2dm"))

tab <- kable(mask, booktabs = TRUE, linesep = "",
             caption = "Regex masks for strata")
# %>%
  # footnote(general = footnote, general_title = "")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Analysis Plan Matrix

For each comparison within each study, we will execute the 7 `CohortMethod` analysis plans in Table \@ref(tab:plan-matrix).

```{r, plan-matrix, echo=FALSE, warning=FALSE}
design <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, file = "
  ID; PS-covariates; PS-adjustment; Outcome model; Risk-window end
  1; ; ; Cox; target exposure end
  2; FE defaults; variable match; Cox (conditional); target exposure end
  3; FE defaults; 5 strata; Cox (conditional); target exposure end
  4; ; ; Cox; observational period end
  5; FE defaults; variable match; Cox (conditional); observational period end
  6; FE defaults; 5 strata; Cox (conditional); observational period end
  7; ; ; Cox; target exposure end or escalation
  8; FE defaults; variable match; Cox (conditional); target exposure end or escalation
  9; FE defaults; 5 strata; Cox (conditional); target exposure end or escalation
")

footnote <- c("FE: FeatureExtraction package (365 days long-term, 180 days medium-term and 30 days short-term look-back)",
              "PS: Propensity score")

tab <- kable(design, linesep = "", booktabs = TRUE,
             caption = "CohortMethod analysis descriptions") %>%
  footnote(general = footnote, general_title = "")


if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped", font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

## Configuration files

The LEGEND-T2DM studies depend on several configuration files.
We directly provide some of these files, while others we generate through executable `R` scripts before distributing study packages.

### Executable R scripts

* `R/CreateAnalysisDetails.R`
* `extra/GenerateExposureCohortDefinitions.R`

### File dependencies

#### For all LEGEND-T2DM studies

* `baseCohort.json` : Base cohort object that we programmatically manipulate to generate all `class` and `drug` exposure cohorts

* `inst/settings/ExposuresOfInterest.csv` : List of `RxNorm` concepts for individual `drug` ingredient and `Drug class` specifications for all exposure cohorts.  File columns:
  - `type` : `Drug` for individual ingredients or `Drug class` for ingredient classes
  - `name` : Unsure if currently used; longer name
  - `cohortId` : Unique concept ID used to make cohort ID
  - `conceptId` : `RxNorm` concept ID if individual ingredient, otherwise unique `Drug class` ID  in `c(1, 2, 3, 4)`
  - `class` : Reference to `Drug class` `conceptId` if individual ingredient, otherwise `NA`
  - `shortName` :  Unsure if currently used; short name
  - `order` : Unsure if currently used; order to display
  - `includedConceptIds` : RxNorm concept IDs if `Drug class`, otherwise `NA`

* `inst/settings/OutcomesOfInterest.csv` : List of outcome cohorts to import into study package and instantiate during execution.  File columns:
  - `cohortId` : Unique cohort ID inside study package
  - `atlasId` : Unique cohort ID on JnJ ATLAS instance (currently) for importing into package
  - `atlasName` : Short name for `Cohort Diagnostics`
  - `name` : File prefix name for `.json` and `.sql` specification
  - `description` : Human-readable, short description of cohort
  - `cite` : List of `Rmd` citation keys
  - `isNew` : Indicate a new cohort for LEGEND-T2DM

* `inst/settings/Indications.csv` : List of study-specific settings.  File columns:
  - `indicationId` : Unique study ID key, e.g., `class`
  - `indicationName` : Study name, e.g., `Class-vs-class`
  - `filterConceptIds` : Concept IDs to remove from propensity score models
  - `positiveControlIdOffset` : Offset with which to start numbering positive control outcomes

#### For `class` studies

The following files are generated by `extra/GenerateExposureCohortDefinitions.R`:

* `inst/settings/classCohortsToCreate.csv` : List of exposure cohorts to create, including both main and `heterogeneity` study-stratified cohorts.  File columns:
  - `atlasId` : Unique cohort ID
  - `atlasName` : Short name that `CohortDiagnostics` will display
  - `cohortId` : Must == `atlasId` (unsure if used)
  - `name` : File prefix name for `.json` and `.sql` specification

* `inst/settings/classTcosOfInterest.csv` : List of TC pairs to compare in the `class` study.  File name includes `Tco` and some columns for historical reasons; all outcomes are executed by default.  File columns :
  - `targetId` : Target cohort ID
  - `comparatorId` : Comparator cohort ID
  - `outcomeIds` : Currently unused; defaults to `-1`
  - `excludedCovariateConceptIds` : Currently unused; defaults to `NA`
  - `targetName` : Unsure if currently used; short name for target cohort
  - `comparatorName` Unsure if currently used; short name for comparator cohort

### Table specification files

The following files specify the diagnostics and results data models on the public Postgresql server:

  * `inst/settings/PsAssessmentModelSpecs.csv` : Propensity score assessment additions to the `legendt2dm_*_diagnostics` schema

  * `inst/settings/ResultsModelSpecs.csv` : All study results for the `legendt2dm_*_results` schema

These files serve as arguments to `createDataModelSqlFile` to generate

* `inst/sql/postgresql/CreatePsAssessmentTables.sql`

* `inst/sql/postgresql/CreateResultsTables.sql`

and to `uploadResultsToDatabase` to upload diagnostics and results to the public Postgresql server.

## `CohortMethod` function arguments

```{r, test, code=xfun::read_utf8("../../R/CreateAnalysisDetails.R")}
```

# Data models

This document describes the data models for storing the output of the `CohortDiagnostics` and `CohortMethod` study artifacts for the LEGEND-T2DM study.
Table \@ref(tab:schema) provides the `schema` names.

```{r schema, echo=FALSE, warning=FALSE}
schema <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, file = "
  legendt2dm_class_diagnostics ; legendt2dm_class_results
  legendt2dm_drug_diagnostics ; legendt2dm_drug_results
  legendt2dm_outcome_diagnostics
")

tab <- kable(schema, col.names = c("Cohort diagnostics", "Study results"),
             linesep = "", booktabs = TRUE,
             caption = "LEGEND-T2DM public study schema")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

## Fields with minimum values

Some fields contain patient counts or fractions that are easily converted to patient counts.
To prevent identifiability, these fields are subject to a minimum value.
When the value falls below this minimum, it is replaced with the negative value of the minimum.
For example, if the minimum subject count is 5, and the actual count is 2, the value stored in the data model will be -5, which could be represented as '\<5' to the user.
Note that the value 0 is permissible, as it identifies no persons.

## Cohort diagnostics data model

```{r echo=FALSE, results="asis", warning=FALSE, message=FALSE}
specifications <- readr::read_csv(system.file("/settings/resultsDataModelSpecification.csv", package = "CohortDiagnostics")) %>%
  filter(optional == "No")

tables <- split(specifications, specifications$tableName)

for (table in tables) {
  header <- sprintf("### Table %s", table$tableName[1])

  table <- table %>%
    select(Field = .data$fieldName, Type = .data$type, Key = .data$primaryKey
           # , Description = .data$description
           ) %>%
    kable(linesep = "", booktabs = TRUE, longtable = TRUE)

  if (knitr::is_latex_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(latex_options = "striped", font_size = latex_table_font_size) %>%
                 column_spec(1, width = "10em") %>%
                 column_spec(2, width = "5em") %>%
                 column_spec(3, width = "3em") %>%
                 column_spec(4, width = "16em"))
  } else if (knitr::is_html_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(bootstrap_options = "striped"))
  }
}
```

## PS assessment data model

```{r echo=FALSE, results="asis", warning=FALSE, message=FALSE}
specifications <- readr::read_csv(system.file("settings/PsAssessmentModelSpecs.csv", package = "LegendT2dm"))

tables <- split(specifications, specifications$tableName)

for (table in tables) {
  header <- sprintf("### Table %s", table$tableName[1])

  table <- table %>%
    select(Field = .data$fieldName, Type = .data$type, Key = .data$primaryKey,
           Description = .data$description
           ) %>%
    kable(linesep = "", booktabs = TRUE, longtable = TRUE)

  if (knitr::is_latex_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(latex_options = "striped", font_size = latex_table_font_size) %>%
                 column_spec(1, width = "10em") %>%
                 column_spec(2, width = "5em") %>%
                 column_spec(3, width = "3em") %>%
                 column_spec(4, width = "16em"))
  } else if (knitr::is_html_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(bootstrap_options = "striped"))
  }
}
```

## Study results data model

```{r echo=FALSE, results="asis", warning=FALSE, message=FALSE}
specifications <- readr::read_csv(system.file("settings/ResultsModelSpecs.csv", package = "LegendT2dm"))

tables <- split(specifications, specifications$tableName)

for (table in tables) {
  header <- sprintf("### Table %s", table$tableName[1])

  table <- table %>%
    select(Field = .data$fieldName, Type = .data$type, Key = .data$primaryKey,
           Description = .data$description
           ) %>%
    kable(linesep = "", booktabs = TRUE, longtable = TRUE)

  if (knitr::is_latex_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(latex_options = "striped", font_size = latex_table_font_size) %>%
                 column_spec(1, width = "10em") %>%
                 column_spec(2, width = "5em") %>%
                 column_spec(3, width = "3em") %>%
                 column_spec(4, width = "16em"))
  } else if (knitr::is_html_output()) {
    writeLines("")
    writeLines(header)

    writeLines(table %>%
                 kable_styling(bootstrap_options = "striped"))
  }
}
```

# Database access

All `LegendT2dm` tables are located on OHDSI's public `PostgreSQL` server that is the backend for the apps on (https://data.ohdsi.org).
Study PIs can provide read-only credentials that can be stored encrypted on individual computers.
For this latter task, we recommend using the cross-platform `keyring` `R` package to securely store: `legendt2dmServer`, `legendt2dmDatabase`, `legendt2dmUser`, `legendt2dmPassword` and `legendT2dmDriverPath` and specifying connection details via:

```{r, eval=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = keyring::key_get("legendT2dmDriverPath"))

legendT2dmConnectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = paste(keyring::key_get("legendt2dmServer"),
                 keyring::key_get("legendt2dmDatabase"),
                 sep = "/"),
  user = keyring::key_get("legendt2dmUser"),
  password = keyring::key_get("legendt2dmPassword"))
```

# R Shiny

## Exposure and outcome cohort characteristics and diagnostics

To explore cohorts and their diagnostics through an R Shiny application, we use the `LegendT2dmCohortExplorer` package.
The following code will load all `class` cohorts:

```{r, eval=FALSE}
LegendT2dmCohortExplorer::launchCohortExplorer(cohorts = "class",
                                               connectionDetails = legendT2dmConnectionDetails)
```

Use `cohorts = "drug"` and `cohorts = "outcome"` to explore the Drug-vs-Drug Study exposure and outcome cohorts, respectively.
This Shiny applications is based on `CohortDiagnostics v2.1` and it is also possible to explore cohorts using local files by specifying the `dataFolder` argument.

A useful `Cohort name` filter is `main(` (see Table \@ref(tab:category-codes)).

## Comparative effectivess and safety evidence

To explore the comparative effectiveness and safety evidence through an R Shiny application, we use the `LegendT2dmEvidenceExplorer` package.
The following code will load all `class` cohort comparison evidence:

```{r, eval=FALSE}
LegendT2dmEvidenceExplorer::launchEvidenceExplorer(cohorts = "class",
                                                   connectionDetails = legendT2dmConnectionDetails)
```

Use `cohorts = "drug"` to explore the `drug` cohort comparison evidence.
It is also possible explore evidence using local files by specifying the `dataFolder` argument.

# Definitions

## Orphan concepts

Orphan concepts are (source) concepts that likely should be included in a concept set, but aren't, often because of vocabulary issues.
This is a rough outline of the process:

* Given a concept set expression, find all included concepts.
* Find all names of those concepts, including synonyms, and the names of source concepts that map to them.
* Search for concepts (standard and source) that contain any of those names as substring.
* Filter those concepts to those that are not in the original set of concepts (i.e. orphans).
* Restrict the set of orphan concepts to those that appear in the CDM database as either source concept or standard concept.

# HOWTOs

## Render human-readable cohort definition

The `LegendT2dm` provides functions to render cohort definitions into `Rmarkdown` for output to PDF or HTML.
Here is an example of rendering the complete definition for new-users of SGLT2 inhibitors:

```{r, render_cohort, results="asis", eval=FALSE}
cohortJson <- SqlRender::readSql(system.file("cohorts", "class", "ID301100000.json",
                                             package = "LegendT2dm"))

LegendT2dm::printCohortDefinitionFromNameAndJson(
  name = "Class-vs-Class Exposure (SGLT2 New-User) Cohort / OT1",
  json = cohortJson)
```

# Dissemination

## Publications

If possible, LEGEND-T2DM publications should be drafted in `Rmarkdown` with `*.Rmd` source files version-controlled on `github`.
Initial drafting can begin in a private repository, but final versions should end up in `ohdsi-studies/LegendT2dm/Documents`.
Currently, we have the following documents:

* `Protocol/Protocol.Rmd`: official protocol registered (soon) with EU PASS
* `Protocol/MedRvix.Rmd`: reformatted protocol for `medRvix`, available at: (add url)
* `Protocol/BmjOpen.Rmd`: reformatted protocol for BMJOpen (to be submitted)

## Presentations

Nothing yet.
We have a cultural war here between `PowerPoint`, `latex` and `Rmarkdown`.
