---
title: "| RESEARCH PROTOCOL\n| \n| Large-scale evidence generation and evaluation
  across a network of databases for type 2 diabetes mellitus\n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    toc: false
    number_sections: yes
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
- \renewcommand{\and}{\\}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: ''
  version: 1.0.0
# subtitle: 'Version: `r params$version`'
link-citations: yes
csl: jamia.csl
author: |
  Rohan Khera MD MS\ $^{\tiny{\text{1,2}}}$ \and
  Martijn J Scheumie PhD\ $^{\tiny{\text{3,4}}}$ \and
  Yuan Lu ScD\ $^{\tiny{\text{1,2}}}$ \and
  Anna Ostropolets MD\ $^{\tiny{\text{5}}}$ \and
  Ruijun Chen MD\ $^{\tiny{\text{6}}}$  \and
  George Hripcsak MD MS\ $^{\tiny{\text{5,7}}}$ \and
  Patrick B Ryan PhD\ $^{\tiny{\text{3,5}}}$, \and
  Harlan M Krumholz MD SM\ $^{\tiny{\text{1,2}}}$ \and
  Marc A Suchard MD PhD\ $^{\tiny{\text{4,8,9,10}}}$
---

```{r setup, include=FALSE}
library(LegendT2dm)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(dplyr)
options(knitr.kable.NA = "")
options(knitr.table.format = function() {
  if (knitr::is_latex_output()) {
    "latex"
  } else if (knitr::is_html_output()) {
    "html"
  } else {
    "pipe"
  }
})

latex_table_font_size <- 8
```

\footnotesize

1. Section of Cardiovascular Medicine, Department of Internal Medicine, Yale School of Medicine, New Haven, CT
2. Center for Outcomes Research and Evaluation, Yale-New Haven Hospital, New Haven, CT
3. Epidemiology Analytics, Janssen Research and Development, Titusville, NJ
4. Department of Biostatistics, UCLA Fielding School of Public Health, University of California, Los Angeles, CA
5. Department of Biomedical Informatics, Columbia University Medical Center, New York, NY
6. Department of Translational Data Science and Informatics, Geisinger, Danville, PA
7. New York-Presbyterian Hospital, New York, NY
8. Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles, CA
9. Department of Human Genetics, David Geffen School of Medicine at UCLA, University of California, Los Angeles, CA
10. VA Informatics and Computing Infrastructure, Department of Veterans Affairs, Salt Lake City, UT

\clearpage
\normalsize

## Disclosures {-}

```{r child = "Disclosures.Rmd"}
```

# List of Abbreviations {-}

```{r child = "Abbreviations.Rmd"}
```

\clearpage

# Abstract {-}

```{r child = "ShorterAbstract.Rmd"}
```



```{r child = "MainText.Rmd"}
```

# References {-}

<div id="refs"></div>

\clearpage

\centerline{\Huge Appendix}

# (APPENDIX) Appendix {-}

```{r child = "Appendix.Rmd"}
```
