---
title: "| RESEARCH PROTOCOL\n| \n| Large-scale evidence generation and evaluation
  across a network of databases for type 2 diabetes mellitus\n"
fontsize: 12pt
geometry: margin=1in
output:
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    toc: false
    number_sections: yes
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
- \renewcommand{\and}{\\}
longtable: yes
mainfont: Arial
bibliography: Protocol.bib
params:
  date: '2021-07-26'
  version: 1.0.0
# subtitle: 'Version: `r params$version`'
link-citations: yes
csl: jamia.csl
author: "Rohan Khera MD MS$^{1,2}$, Martijn J Scheumie PhD,$^{3}$ Yuan Lu ScD,$^{1,2}$, Anna Ostropolets MD,$^{4}$ Fan Bu PhD,$^{5}$  Akihiko (Aki) Nishimura,$^{6}$, Ruijun Chen MD,$^{7}$  George Hripcsak MD MS,$^{4,8}$  Patrick B Ryan PhD,$^{3,4}$, Harlan M Krumholz MD SM,$^{1,2}$ Marc A Suchard MD PhD$^{5,9}$"
---

```{r setup, include=FALSE}
library(LegendT2dm)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(dplyr)
options(knitr.kable.NA = "")
options(knitr.table.format = function() {
  if (knitr::is_latex_output()) {
    "latex"
  } else if (knitr::is_html_output()) {
    "html"
  } else {
    "pipe"
  }
})

latex_table_font_size <- 8
```

$^{1}$ Section of Cardiovascular Medicine, Department of Internal Medicine, Yale School of Medicine, New Haven, CT
$^{2}$ Center for Outcomes Research and Evaluation, Yale-New Haven Hospital, New Haven, CT
$^{3}$ Epidemiology Analytics, Janssen Research and Development, Titusville, NJ
$^{4}$ Department of Biomedical Informatics, Columbia University Medical Center, New York, NY
$^{5}$ Department of Biomathematics, David Geffen School of Medicine at UCLA, University of California, Los Angeles, CA
$^{6}$ Department of Biostatistics, Johns Hopkins University, Baltimore, MD
$^{7}$ Department of Translational Data Science and Informatics, Geisinger, Danville, PA
$^{8}$ NewYork-Presbyterian Hospital, New York, NY
$^{9}$ Department of Biostatistics, Fielding School of Public Health, University of California, Los Angeles, CA

## Disclosures {-}

```{r child = "Disclosures.Rmd"}
```

\clearpage

# List of Abbreviations {-}

```{r child = "Abbreviations.Rmd"}
```

# Abstract {-}

```{r child = "ShorterAbstract.Rmd"}
```



```{r child = "MainText.Rmd"}
```

# References {-}

<div id="refs"></div>

# (APPENDIX) Appendix {-}

```{r child = "Appendix.Rmd"}
```
